---
title: "Chapter 12 Week 4"
author: "Marissa"
date: "2/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Chapter 12 Tidy Data Notes

#### 12.1.1 Prerequisites
```{r}
library(tidyverse)
```
### 12.2 Tidy Data
```{r}
# runs the tables
table1

table2

table3

table4a 

table4b 

```

Dataset Tidy:
Each variable must have its own column
Each observation must have its own row
Each value must have its own cell

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
```

## 12.2.1 Exercises
```{r}
#1 Using prose, describe how the variables and observations are organised in each of the sample tables.

#Columns are their own variables, meaning year only has year or country only has a country.
table1

# Combined columns, such as cases and population in the type column.
table2

# Rate is a combine column of separated by a /.
table3

# Contains the cases, with the the years separated into separate columns.
table4a 

# Contains the population.
table4b 

```

```{r}
#2 Compute the rate for table2, and table 4a+table 4b. 
  #1 Extract the bumber of TB case per country per year.
  #2 Extract the matching population per country per year.
  #3 Divide the cases by population and mulitply by 1000
  #4 Sore back in the appropraite place
# Which representation is easiest to work with? Which is hardest? Why?

#Table2
cases <- table2 %>%
  filter(type == "cases")
populations <- table2 %>%
  filter(type == "population")
merged <- cbind(cases, populations)
merged$rates= (cases$count / populations$count)*1000
merged

#Tables4
rates_99 <- (table4a$`1999` / table4b$`1999`)*1000
rates_00 <- (table4a$`2000`/ table4b$`2000`)*1000
new <- table4a['country']
new$rates_00 = rates_00
new$rates_99 = rates_99
new

#Table 4a and 4b were easier to work with, whereas Table 2 had to be separated and filter before doing any of the calculations.
```


### 12.3 Pivoting

#### 12.3.1 Longer
A common problem is a dataset where some of the column names are not names of variables, but values of a variable.

pivot_longer() makes wide tables narrower and longer
```{r}
table4a
# Changes 1999 to year and 2000 to cases
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

```{r}
table4b
# Tidying table 4b by using pivot_longer
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

```{r}
# Combines the tidied table4a and table4b
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```

#### 12.3.2 Wider
pivot_wider() is the opposite of pivot_longer(). You use it when an observation is scattered across multiple rows.

pivot_wider() makes long tables shorter and wider.
```{r}
table2
# separates type into two columns of case and population with the count as the values
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```

## 12.3.3 Exercises
```{r}
#1 Why are pivot_longer and pivot_wider not perfectly symmetrical?
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
# Since the shape of the original tibble has changed and the orignal placement of columns and values. Pivot_wider() creates each year as it's own column instead of a double value column.
```

```{r}
#2 Why does this code fail?

#table4a %>% 
#  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")

# Fails because we tried to use column names as integers
```

```{r}
#3 Why would happen if you widen this table? Why? How could you add a new column to uniquely identify each value?
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
# Receive an error when trying to widen the table, due to the duplication of names. To avoid the error, I would create columns for age and one for height.
people %>%
  pivot_wider(names_from = names, values_from = values)

```

```{r}
#4 Tidy the simple tibble below. Do you need to make it wider or longer? What are the variables?
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg%>%
  pivot_longer(c('male','female'), names_to="sex",values_to=)
```

### 12.4 Separating and Uniting

#### 12.4.1 Separate
separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears.
```{r}
table3
# separates rate into cases and population at the forward slash
table3 %>% 
  separate(rate, into = c("cases", "population"))
# could also write as:
#table3 %>% 
  #separate(rate, into = c("cases", "population"), sep = "/")
```

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

#### 12.4.2 Unite
unite() is the inverse of separate(): it combines multiple columns into a single column
```{r}
table5 %>% 
  unite(new, century, year)
```

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

## 12.4.3 Exercises
```{r}
#1 What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.

# The extra argument 'g'. The extra tells what to do if there are too many
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

# The fill argument 'f'. The fill argument tells it what to do if there is something missing.
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
```

```{r}
#2 Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?

#The remove argument removes/discards the columns used to either unite or separate. Setting it to FALSE, keeps the original columns.
```

```{r}
#3 Compare and contrast separate() and extract(). Why are there three variations of separation (position, separator, groups), but only one unite?

# Separate() splits a column into multiple columns.
# Extract() splits a column into multuple columns as well.
# Separate and Extract both start with a single column broken into several columns.
# Unite takes multiple columns converts it into a single column.
```

### 12.5 Missing Values
* Explicitly, i.e. flagged with NA.
* Implicitly, i.e. simply not present in the data.
```{r}
# Missing two values
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

```{r}
stocks %>% 
  complete(year, qtr)
```

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
```

```{r}
# Fill takes a set of columns where you want missing values to be replaced by the most recent non-missing value (sometimes called last observation carried forward).
treatment %>% 
  fill(person)
```

## 12.5.1 Exercises
```{r}
#1 Compare and contrast fill arguments to pivot_wider() and complete().

treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

# Carries Derrick Whitmore to fill in the NA in person
treatment %>% 
  fill(person)

# Creates an error there is missing data from the person column. Must do pivot_longer() as well to drop the NA values.

#pivot_wider(names_from = person, values_from = response)

# Finds unique values and creates several NA in person
treatment %>%
  complete(treatment, response)
```

```{r}
#2 What does the direction argument to fill() do?

# The direction tells it in which direction it should take the last data to fill in the blanks with.


treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

# Carries Katherine Burke to fill in the NA
fill(treatment, person, .direction = "up")
```

### 12.6 Case Study
```{r}
who
```

```{r}
# Focus on the values that are present and create new colum names.
who1 <- who %>% 
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  )
who1
```

```{r}
# Structure of the values
who1 %>% 
  count(key)
```
* First three letters of each column denote new or old cases of TB.
* Next two letters describe the type of TB:
  + rel = relapse
  + ep = extrapulmonary TB
  + sn = smear negative
  + sp = smear positive
* Sixth letter gives the sex of the patient. M = male; F = female
* Numbers represent the age:
  + 014 = 0 – 14 years old
  + 1524 = 15 – 24 years old
  + 2534 = 25 – 34 years old
  + 3544 = 35 – 44 years old
  + 4554 = 45 – 54 years old
  + 5564 = 55 – 64 years old
  + 65 = 65 or older

```{r}
# Fixes variable names that are inconsistent
who2 <- who1 %>% 
  mutate(names_from = stringr::str_replace(key, "newrel", "new_rel"))
who2
```

```{r}
# Separates at each underscore
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")

who3
```

```{r}
who3 %>% 
  count(new)
```

```{r}
# Drops columns
who4 <- who3 %>% 
  select(-new, -iso2, -iso3)
```

```{r}
# Separates sexage into sex and age
who5 <- who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)

who5
```

```{r}
# TIDY DATASET
who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  ) %>% 
  mutate(
    key = stringr::str_replace(key, "newrel", "new_rel")
  ) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

## 12.6.1 Exercises
```{r}
#1 In this case study I set values_drop_na = TRUE just to make it easier to check that we had the correct values. Is this reasonable? Think about how missing values are represented in this dataset. Are there implicit missing values? What's the difference between an NA and 0?

# values_drop_na = TRUE may be reasonable depending on no cases of TB were represented. If there are no 0s in the dataset, then NA may represent no cases of TB. Therefore dropping the NA values skews the data. 0 is an integer while NA means a blank in the dataset.
```

```{r}
#2 What happens if you neglect the mutate() step?
#(mutate(names_from = stringr::str_replace(key, "newrel", "new_rel")))

# The mutate steps replaces the variable name with new_rel for consistency purposes and to avoid following errors. It allows for the separation at the '_' in the following step to avoid multiple columns.
```

```{r}
#3 I claimed that iso2 and iso3 were redundant with country. Confirm this claim.

select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)
```

```{r}
#4 For each country, year, and sex compute the total number of cases of TB.

who5 %>%
  group_by(country, year, sex) %>%
  summarise(cases = sum(cases))

```

### 12.7 Non-Tidy Data
Two main reasons to use other data structures:
Alternative representations may have substantial performance or space advantages.
Specialised fields have evolved their own conventions for storing data that may be quite different to the conventions of tidy data.
