---
title: "Chapter 12 Week 4"
author: "Marissa"
date: "2/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```
# 12.2 Tidy Data
```{r}
# runs the tables
table1

table2

table3

table4a 

table4b 

```

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
```

## 12.2.1 Exercises
1.
2.

# 12.3 Pivoting

## 12.3.1 Longer
A common problem is a dataset where some of the column names are not names of variables, but values of a variable.

pivot_longer() makes wide tables narrower and longer
```{r}
table4a
# Changes 1999 to year and 2000 to cases
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

```{r}
table4b
# Tidying table 4b by using pivot_longer
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

```{r}
# Combines the tidied table4a and table4b
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```

## 12.3.2 Wider
pivot_wider() is the opposite of pivot_longer(). You use it when an observation is scattered across multiple rows.

pivot_wider() makes long tables shorter and wider.
```{r}
table2
# separates type into two columns of case and population with the count as the values
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```

## 12.3.3 Exercises
1.
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

2.
```{r}
#table4a %>% 
 # pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
```

3.
```{r}
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

4.
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

# 12.4 Separating and Uniting

## 12.4.1 Separate
separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears.
```{r}
table3
# separates rate into cases and population at the forward slash
table3 %>% 
  separate(rate, into = c("cases", "population"))
# could also write as:
#table3 %>% 
  #separate(rate, into = c("cases", "population"), sep = "/")
```

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

## 12.4.2 Unite





